market-justice-preferences

### 2024-05-08 paper.tex
update plots and tables antecedentes y metodos

Since its origins, educational institutions have been related to the
idea of social mobility and access to better opportunities. Therefore,
the consistent evidence of the high level of social reproduction at the
school level represents a threat to the promise of education and a
meritocratic system (\citeproc{ref-bourdieu_reproduction_1990}{Bourdieu
and Passeron 1990}). A large part of the research in this field at an
international level has addressed the extent to which the social origin
of students affects their academic results and their life opportunities
(\citeproc{ref-vonhippel_test_2019}{Von Hippel and Hamrock 2019}),
confirming that schools have severe difficulties in closing the gaps of
origin. Besides this socioeconomic perspective on school opportunities,
recent research has addressed to what extent inequalities in the school
context are also influencing students' perceptions, beliefs, and
attitudes: Are social inequalities perceived at the school context? Are
they rejected by the students, particularly those who are worst-off in
socioeconomic terms? Or, Is there evidence at the school level that
social inequalities are tolerated and even justified?
(\citeproc{ref-batruch_belief_2022}{Batruch et al. 2022};
\citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al. 2015}).~

In the present paper, we deal with the justification of social
inequalities by eighth-grade students in Chile, a country characterized
by a highly stratified educational system. In particular, we focus on
market justice preferences (Lane, 1986), which refer to the preferences
for distributing public goods (as health and education) based on
criteria such as competition and payment capacity. Although from a
rational point of view it could be expected an opposition to market
justice by the underprivileged majority,~ we argue that in a social
environment characterized by a promotion of meritocratic ideals - as the
school system - would lead to the opposite: a larger market justice
preferences.~

Given that the school environment has an important focus on performance,
achievement and acknowledgment, meritocracy has been one of the
principal concepts used for understanding and even for justifying
performance differences among students. Meritocracy is a distributive
system based on the belief that people should be rewarded and promoted
based on their abilities, knowledge, and achievements
(\citeproc{ref-young_rise_1958}{Young 1958}). It is often seen as a way
to create equal opportunities and fairness, as individuals can rise to
positions of power and influence based on their own merit rather than
their background or connections. However, some argue that meritocracy
can actually lead to tolerating or even justifying social inequalities,
as it can create a hierarchy where those who already have resources and
advantages are more likely to succeed. In this regard, a great deal of
academic research about meritocracy delves into the assessment of to
what extent rewards and privileges in society are related to merit,
emphasizing the so-called unfulfillable promise of meritocracy
(\citeproc{ref-mijs_stratified_2016}{J. Mijs 2016}). Complementing this
agenda, a second and emerging research area deals with subjective
aspects of meritocracy, such as perceptions and beliefs.~

@@

The perception of meritocracy refers to how individuals view and
understand the concept of meritocracy in their own society
(\citeproc{ref-duru-bellat_who_2012}{Duru-Bellat and Tenret 2012};
\citeproc{ref-castillo_meritocracia_2019}{Juan C. Castillo et al.
2019}). This perception can vary greatly depending on individual
experiences, social, economic, and cultural background. Some people may
see meritocracy as a fair and just system that allows anyone to succeed
based on their abilities and hard work. In contrast, others may view it
as a myth or a cover for existing power dynamics and inequality, serving
to maintain and even reinforce inequality
(\citeproc{ref-lampert_meritocratic_2013}{Lampert 2013};
\citeproc{ref-mijs_paradox_2021}{J. J. B. Mijs 2021}). Based in this
last perspective, we argue that individuals with a higher perception of
meritocracy will show a larger justification of social inequalities, as
individual achievement would be seen as rewarded and social policies as
less necessary (\citeproc{ref-batruch_belief_2022}{Batruch et al.
2022}).

@@

The present paper deals with the association between the perception of
meritocracy and the justification of social inequalities, with two main
focuses. Firstly, it assesses the justification of social inequalities
in social policy domains, such as health, pensions, and education. We
argue that individuals who perceive more meritocracy would be more
willing to justify better services in these domains for those with
higher incomes.~ Secondly, we focus on the student-age population as we
point out that it is possible to track down the origin of meritocratic
beliefs (and their consequences) to early socialization processes. To
this regard, we take into account the family and the school as two main
socialization agencies that play a significant role in the socialization
of cultural beliefs by transmitting cultural norms, values, and
expectations to young people.~

@@

Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why
(\citeproc{ref-kluegel_beliefs_1987}{Kluegel and Smith 1987}),
highlights that people's explanations and justifications of social
inequality are closely tied to their judgments of deservingness. The
influence of ideologies (\citeproc{ref-wegener_dominant_1995}{Wegener
and Liebig 1995}) and cultural schemas
(\citeproc{ref-homan_being_2017}{Homan, Valentino, and Weed 2017}) is
pivotal in shaping these explanations by offering symbolic
representations that frame societal structures and expectations. While
significant attention has been paid to wage inequality, income
distribution, and payment differentials in the literature {[}Juan Carlos
Castillo (\citeproc{ref-castillo_legitimacy_2011}{2011}); {]} (Castillo,
2011; Evans et al., 2010; Jasso, 1999; Shariff et al., 2016),~ there has
been less examination of public beliefs about which life domains should
be governed by market relations and even less about children's
acceptance or rejection of these market principles. This oversight is
notable given the extensive encroachment of market logic into public
goods, welfare policy, and social services over the past five decades
(Centeno \& Cohen, 2012; Harvey, 2007), affecting areas such as
pensions, health services, and education. The justification of social
inequality based on marketype criteria has been conceptualized as the
individuals' adherence to one specific justice evaluation, the market
justice, that is, affording legitimacy to the allocation of goods and
services based on prices and individuals' ability to pay (Boltanski \&
Chiapello- 2007; Lane 1986; Streeck 2012).

Robert E. Lane proposed the underpinnings of market justice, which he
differentiated from political justice. For him, ``it is the genius of
the market to stimulate wants without at the same time stimulating a
sense of deserving more than one gets'' (1986, p.~384). Following the
theory of relative deprivation --a social phenomenon arising when
individuals cannot afford what most others in their environment can
(Merton 1950)-- Lane notes that, in market settings, social comparisons
are more likely to motivate increased effort rather than feelings of
acute injustice because individuals attribute outcomes to their actions.
Although empirical research has shown that, contrary to Lane's
observation, relative deprivation, even in market settings, produces
feelings of dissatisfaction, anger, and resentment that might motivate
forms of collective action such as protests and revolt (Greitemeyer and
Sagioglou 2016; Mishra and Carleton 2015; Smith et al.~2012; Power 2018;
Power, Madsen, and Morton 2020), the conceptualization of market justice
has been since then closely coupled to the merit principle for
allocating outcomes: the claim that the unequal levels of well-being
individuals enjoy ought to be, to some extent, a function of their
talents and efforts, regardless of their needs or membership, the two
latter being the realm of political justice and its closely coupled
principles of need --allocating outcomes to those who require them
most-- and equality --allocating the same outcome to everyone--, which
have been at the center of welfarism (see Wilson 2003). The underlying
notion of market justice also resonates in its application to welfare
regimes, providing a framework for understanding the varied global
approaches to managing social services.

The management of social services manifests in varying approaches across
nations, with substantial differences in funding and delivery methods
(Jensen, 2008; Stoy, 2014). Nordic countries, for example, predominantly
employ public agencies to produce and provide social services, funding
these through collective taxation and offering them in kind to the
majority of citizens. This system prioritizes political justice, placing
it above market mechanisms in accessing services. In contrast, other
countries rely more heavily on for-profit entities and private funding,
where service distribution depends mainly on individual financial
capacity of paying user fees, highlighting the influence of market
justice in service allocation. The trend toward marketization of welfare
services has been growing since the 1980s (Salamon, 1993), and this
shift is increasingly evident even in countries where market solutions
have traditionally had a minor role in social policy (Sivesind, 2017).

The question arises whether adults and children justify unequal access
to welfare services based on market justice principles. Influenced by
theories of policy feedback, which suggest that social welfare policies
can reinforce (positive feedback) or undermine (negative feedback)
previous policy trajectories
(\citeproc{ref-fernandez_positive_2013}{Fernandez and Jaime-Castillo
2013}) (Fernandez and Jaime-Castillo 2013; Pierson 2000; Weaver 2010),
citizens' beliefs about market justice are likely also shaped by the
institutional and social contexts they encounter. Indeed, the
justification of inequality in access to essential services like
education and health, based on one's ability to pay, shows significant
variation across countries, as demonstrated by international surveys,
although they do not usually include children in their samples. For
instance, the 2019 International Social Survey Program (ISSP) provides
insights into how adults perceive inequality in accessing welfare
services. Figure 1 illustrates the variation in agreement levels
regarding whether it is just for individuals with higher incomes to
purchase better education or healthcare. Notably, in 18 of the 29
countries analyzed, there is a greater justification for market
inequality in healthcare access than in education. Nevertheless, the
general sentiment typically ranges from `somewhat unjust' to 'neither
just nor unjust, with mixed feelings.

While contemporary policy developments have increasingly embraced the
marketization of social welfare in areas such as education, pensions,
and health services, questions about public acceptance of these changes
persist. Lindh (\citeproc{ref-lindh_public_2015}{2015}) analysis of ISSP
2009 data from 17 OECD countries reveals a general lack of support for
market-based distribution of social services, suggesting widespread
disapproval of market stratification of essential services. This finding
is corroborated by Soler-Martínez and colleagues' (2023) research from
Latinobarómetro 2020 across 18 Latin American countries, where concerns
about health and education access predominated over income inequality.
These results indicate that reforms toward welfare marketization are
typically driven by elite political decisions rather than grassroots
demand.~

Despite high-income inequality and limited social mobility in Latin
America, there is a prevalent belief that individuals are solely
responsible for their economic outcomes, a view that varies across the
region
(\citeproc{ref-buccaMeritBlameUnequal2016}{\textbf{buccaMeritBlameUnequal2016?}};
\citeproc{ref-chongMysteryDiscriminationLatin2008}{\textbf{chongMysteryDiscriminationLatin2008?}};
\citeproc{ref-torcheIntergenerationalMobilityInequality2014}{\textbf{torcheIntergenerationalMobilityInequality2014?}};
\citeproc{ref-salgadoInequalityStratificationLatin2023}{\textbf{salgadoInequalityStratificationLatin2023?}}).
The reliance on private welfare providers and widespread user fees
(Molyneux, 2008) adds complexity to this context. Yet, research on
children's justification of market-based inequalities in accessing
welfare services remains limited, especially in Latin America,
highlighting a significant gap in understanding how younger generations
view market-based access to welfare and whether these views are
associated with their meritocratic beliefs.~

@@

Empirical research on meritocracy has increased along with the
philosophical-normative discussion on meritocracy in recent years.
Particularly from a sociological perspective, meritocracy has been used
in research on social mobility to characterize societies with low
mobility that threaten the meritocratic ideal
(\citeproc{ref-goldthorpe_myth_2003}{Goldthorpe 2003}). More recently,
sociology and social psychology research has attended to the subjective
aspects vis-a-vis beliefs in meritocracy. The label of beliefs in this
realm covers a series of areas, such as attitudes, perceptions, and
preferences (Castillo et al), whereby most of the link this subjective
dimensions to individual socio-structural factors and~ context-level
determinants. For instance, some studies have analyzed how those with
greater privileges believe more in meritocracy (Reynolds \& Chan 2014),
how greater economic inequality increases meritocratic beliefs
(\citeproc{ref-mijs_paradox_2021a}{\textbf{mijs\_paradox\_2021a?}}), and
how larger inequality affects meritocratic beliefs (Morris et al 2022).
Based on these findings, a research agenda has been reinforced on the
legitimizing role of meritocracy, in line with previous studies using
the concept of a just world (Lerner, Dalbert) and the theory of system
justification (Jost \& Major).

How do meritocratic beliefs legitimize inequalities? Empirical studies
have used experiments and surveys to address this question. For
instance, the evidence suggests that just world beliefs correlate
negatively with support for redistributive compensation systems
(\citeproc{ref-frank_performance_2015}{Frank, Wertenbroch, and Maddux
2015}). Conversely, individuals tend to support redistribution when they
believe that the disadvantaged lack the opportunities to succeed
(\citeproc{ref-evans_strong_2018}{Evans and Kelley 2018}). Almås,
Cappelen, and Tungodden (\citeproc{ref-almas_cutthroat_2020}{2020})
found that in a relatively unequal society (the United States), the
highly educated accept inequality significantly more than the less
educated because they perceive inequality as justifiable owing to
differences in productivity (i.e., merit), whereas in a relatively equal
society (Norway), the less educated accept inequality more, but not
significantly more than the highly educated because meritocratic values
are less prevalent. Barr and Miller
(\citeproc{ref-barr_effect_2020}{2020}) also addressed this triple
interaction between the level of inequality in a society, the individual
level of education, and the perceived origin of the disparity (either by
luck or effort) to determine the extent to which inequality is accepted.
They found that the interaction's mechanism varies depending on the
compared societies. Finally, García-Sánchez et al.
(\citeproc{ref-garcia-sanchez_attitudes_2020}{2020}), using data for 41
countries from the International Social Survey Programme (ISSP), found
that the perceived size of the income gap correlated positively with
support for progressive taxation. Still, this association was weaker
among those who endorsed meritocratic and equal opportunity beliefs. In
the same line, experimental research by Durante \& Putterman (2009)
subjects support less redistribution when the initial distribution is
determined according to task performance.

Research about meritocratic beliefs at school age is rather scarce,
leaving a wide research gap as schools are one of the primary
socialization institutions where achievement based on merit explains
success (Erivwo et al 2021). Nevertheless, it is possible to find some
initial works focused on the area of distributive justice at school that
are closely related to meritocratic beliefs, as the ones by Resh and
Sabbagh (quotes). Using justice in grade obtention as a measure of
distributive justice and meritocracy (Sabbagh et al 2006), they find for
instance that a larger sense of distributive justice about grades is
associated to higher socio-economic status
(\citeproc{ref-resh_sense_2010}{Resh 2010}), have a positive effect on
liberal democratic orientation and on trust in people and in formal
institutions (Resh \& Sabbagh 2013), and tend to refrain from violence
and to engage to a greater extent in extra-curricular school activity
and community volunteering (\citeproc{ref-resh_sense_2017}{Resh and
Sabbagh 2017}).

@@

Research indicates various socialization practices at families and
schools during childhood and adolescence that impact dispositional and
behavioral tendencies concerning justifications of inequality in adult
life. The differences in economic understanding across age groups are
consistent with research on cognitive development (Choudhury et al.,
2006). Adolescents with mature socio-cognitive abilities tend to express
stronger preferences for fairness than infants and children (Wynn et
al., 2017). As children grow older, they become more likely to behave
fairly, with their early-emerging strict egalitarianism being replaced
by an increasing endorsement of fairness principles and engagement in
collaborative activities (Huppert et al., 2019; McAuliffe et al., 2017).
In these activities, their fairness views consider individual
contributions, merits, and circumstances (Almås et al., 2010; Huppert et
al., 2019; Sigelman \& Waitzman, 1991). Engelmann and Tomasello (2019)
claim that children's sense of fairness emerges at three years old, and
we can observe it in collaborative activities, where they accept
inequality if the procedure gives everyone an equal chance. Therefore,
children at this age respond to unequal distributions based on
interpersonal concerns, as they already demand equal respect. In any
case, between 3 and 8 years of age, inequitable and anti-meritorious
allocations are evaluated more negatively, but equitable and meritorious
allocations are not evaluated more positively (Elenbaas, 2019).

Some research shows that the social environment in which children
develop, such as family and school, is associated with their prosocial
behaviors by playing an essential role in the transmission of equity
norms (Schunk \& Zipperle, 2023; Kosse et al., 2020). In fact, schools
contribute to institutionalizing and reproducing inequality by promoting
values, norms, practices, and languages familiar to higher-class
families because the dominant group's culture shapes educational
institutions (Bourdieu \& Passeron, 1990). Middle- and upper-class
students are better equipped to face academic challenges and are more
familiar with academic expectations (Mikus et al., 2019). Such
familiarity represents cultural capital in educational contexts because
higher-status students come to school ready to meet these expectations
and reap the benefits (Jack, 2016; Khan, 2011). Conversely, lower-status
children lacking cultural capital must catch up while experiencing
inequitable comparisons (Goudeau \& Croizet, 2017). Additionally,
academic achievement is treated as the outcome of dispositional factors
(e.g., pupils' efforts and talents or lack of them) rather than the
result of differential access to critical resources. Due to the
meritocratic frame schools encourage, both low- and high-status
individuals believe that students' success or failure is not due to
their family background but rather to differences in efforts and talents
(Darnon et al., 2018). In this sense, we believe that the perception of
meritocracy can influence students' judgments about market justice
preferences. Furthermore, we believe there is a difference between
students' perceptions of meritocracy based on their own experience in
school and what they perceive in society at large. Consistently, our
first two hypotheses are:

@@

@@

Table 1 shows the items used, their response categories, and their
frequencies.

>>>>

Since its origins, educational institutions have been related to the
idea of social mobility and access to better opportunities. Therefore,
the consistent evidence of the high level of social reproduction at the
school level represents a threat to the promise of education and a
meritocratic system (\citeproc{ref-bourdieu_reproduction_1990}{Bourdieu
and Passeron 1990}). A large part of the research in this field at an
international level has addressed the extent to which the social origin
of students affects their academic results and their life opportunities
(\citeproc{ref-vonhippel_test_2019}{Von Hippel and Hamrock 2019}),
confirming that schools have severe difficulties in closing the gaps of
origin. Besides this socioeconomic perspective on school opportunities,
recent research has addressed to what extent inequalities in the school
context are also influencing students' perceptions, beliefs, and
attitudes: Are social inequalities perceived at the school context? Are
they rejected by the students, particularly those who are worst-off in
socioeconomic terms? Or, Is there evidence at the school level that
social inequalities are tolerated and even justified?
(\citeproc{ref-batruch_belief_2022}{Batruch et al. 2022};
\citeproc{ref-wiederkehr_belief_2015}{Wiederkehr et al. 2015}).

In the present paper, we deal with the justification of social
inequalities by eighth-grade students in Chile, a country characterized
by a highly stratified educational system. In particular, we focus on
market justice preferences (\citeproc{ref-lane_market_1986}{Lane 1986}),
which refer to the preferences for distributing public goods (as health
and education) based on criteria such as competition and payment
capacity. Although from a rational point of view it could be expected an
opposition to market justice by the underprivileged majority, we argue
that in a social environment characterized by a promotion of
meritocratic ideals - as the school system - would lead to the opposite:
a larger market justice preferences.

Given that the school environment has an important focus on performance,
achievement and acknowledgment, meritocracy has been one of the
principal concepts used for understanding and even for justifying
performance differences among students. Meritocracy is a distributive
system based on the belief that people should be rewarded and promoted
based on their abilities, knowledge, and achievements
(\citeproc{ref-young_rise_1958}{Young 1958}). It is often seen as a way
to create equal opportunities and fairness, as individuals can rise to
positions of power and influence based on their own merit rather than
their background or connections. However, some argue that meritocracy
can actually lead to tolerating or even justifying social inequalities,
as it can create a hierarchy where those who already have resources and
advantages are more likely to succeed. In this regard, a great deal of
academic research about meritocracy delves into the assessment of to
what extent rewards and privileges in society are related to merit,
emphasizing the so-called unfulfillable promise of meritocracy
(\citeproc{ref-mijs_stratified_2016}{Mijs 2016}). Complementing this
agenda, a second and emerging research area deals with subjective
aspects of meritocracy, such as perceptions and beliefs.

@@

The perception of meritocracy refers to how individuals view and
understand the concept of meritocracy in their own society
(\citeproc{ref-duru-bellat_who_2012}{Duru-Bellat and Tenret 2012};
\citeproc{ref-castillo_meritocracia_2019}{Castillo et al. 2019}). This
perception can vary greatly depending on individual experiences, social,
economic, and cultural background. Some people may see meritocracy as a
fair and just system that allows anyone to succeed based on their
abilities and hard work. In contrast, others may view it as a myth or a
cover for existing power dynamics and inequality, serving to maintain
and even reinforce inequality
(\citeproc{ref-lampert_meritocratic_2013}{Lampert 2013};
\citeproc{ref-mijs_paradox_2021}{Mijs 2021}). Based in this last
perspective, we argue that individuals with a higher perception of
meritocracy will show a larger justification of social inequalities, as
individual achievement would be seen as rewarded and social policies as
less necessary (\citeproc{ref-batruch_belief_2022}{Batruch et al.
2022}).

@@

The present paper deals with the association between the perception of
meritocracy and the justification of social inequalities, with two main
focuses. Firstly, it assesses the justification of social inequalities
in social policy domains, such as health, pensions, and education. We
argue that individuals who perceive more meritocracy would be more
willing to justify better services in these domains for those with
higher incomes. Secondly, we focus on the student-age population as we
point out that it is possible to track down the origin of meritocratic
beliefs (and their consequences) to early socialization processes. To
this regard, we take into account the family and the school as two main
socialization agencies that play a significant role in the socialization
of cultural beliefs by transmitting cultural norms, values, and
expectations to young people.

@@

Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why
(\citeproc{ref-kluegel_beliefs_1987}{Kluegel and Smith 1987}),
highlights that people's explanations and justifications of social
inequality are closely tied to their judgments of deservingness. The
influence of ideologies (\citeproc{ref-wegener_dominant_1995}{Wegener
and Liebig 1995}) and cultural schemas
(\citeproc{ref-homan_being_2017}{Homan, Valentino, and Weed 2017}) is
pivotal in shaping these explanations by offering symbolic
representations that frame societal structures and expectations. While
significant attention has been paid to wage inequality, income
distribution, and payment differentials in the literature {[}Castillo
(\citeproc{ref-castillo_legitimacy_2011}{2011}); Evans et al., 2010;
Jasso (\citeproc{ref-jasso_how_1999}{1999}) ; Shariff, Wiwad, and Aknin
(\citeproc{ref-shariff_income_2016}{2016}){]}, there has been less
examination of public beliefs about which life domains should be
governed by market relations and even less about children's acceptance
or rejection of these market principles. This oversight is notable given
the extensive encroachment of market logic into public goods, welfare
policy, and social services over the past five decades
(\citeproc{ref-centeno_arc_2012}{Centeno and Cohen 2012};
\citeproc{ref-harvey_breve_2015}{Harvey 2015}), affecting areas such as
pensions, health services, and education. The justification of social
inequality based on marketype criteria has been conceptualized as the
individuals' adherence to one specific justice evaluation, the market
justice, that is, affording legitimacy to the allocation of goods and
services based on prices and individuals' ability to pay
(\citeproc{ref-boltanski_new_2005}{Boltanski and Chiapello 2005};
\citeproc{ref-lane_market_1986}{Lane 1986};
\citeproc{ref-streeck_citizens_2012}{Streeck 2012}).

Robert E. Lane proposed the underpinnings of market justice, which he
differentiated from political justice. For him, ``it is the genius of
the market to stimulate wants without at the same time stimulating a
sense of deserving more than one gets'' {[}-Lane
(\citeproc{ref-lane_market_1986}{1986}); p.~384{]}. Following the theory
of relative deprivation --a social phenomenon arising when individuals
cannot afford what most others in their environment can (Merton 1950)--
Lane notes that, in market settings, social comparisons are more likely
to motivate increased effort rather than feelings of acute injustice
because individuals attribute outcomes to their actions. Although
empirical research has shown that, contrary to Lane's observation,
relative deprivation, even in market settings, produces feelings of
dissatisfaction, anger, and resentment that might motivate forms of
collective action such as protests and revolt
(\citeproc{ref-greitemeyer_subjective_2016}{Greitemeyer and Sagioglou
2016}; \citeproc{ref-mishra_subjective_2015}{Mishra and Carleton 2015};
\citeproc{ref-smith_relative_2012}{Smith et al. 2012};
\citeproc{ref-power_deprivationprotest_2018}{Séamus A. Power 2018};
\citeproc{ref-power_relative_2020}{Séamus A. Power, Madsen, and Morton
2020}), the conceptualization of market justice has been since then
closely coupled to the merit principle for allocating outcomes: the
claim that the unequal levels of well-being individuals enjoy ought to
be, to some extent, a function of their talents and efforts, regardless
of their needs or membership, the two latter being the realm of
political justice and its closely coupled principles of need
--allocating outcomes to those who require them most-- and equality
--allocating the same outcome to everyone--, which have been at the
center of welfarism (see Wilson 2003). The underlying notion of market
justice also resonates in its application to welfare regimes, providing
a framework for understanding the varied global approaches to managing
social services.

The management of social services manifests in varying approaches across
nations, with substantial differences in funding and delivery methods
(\citeproc{ref-jensen_worlds_2008}{Jensen 2008};
\citeproc{ref-stoy_worlds_2014}{Stoy 2014}). Nordic countries, for
example, predominantly employ public agencies to produce and provide
social services, funding these through collective taxation and offering
them in kind to the majority of citizens. This system prioritizes
political justice, placing it above market mechanisms in accessing
services. In contrast, other countries rely more heavily on for-profit
entities and private funding, where service distribution depends mainly
on individual financial capacity of paying user fees, highlighting the
influence of market justice in service allocation. The trend toward
marketization of welfare services has been growing since the 1980s
(\citeproc{ref-salamon_marketization_1993}{Salamon 1993}), and this
shift is increasingly evident even in countries where market solutions
have traditionally had a minor role in social policy
(\citeproc{ref-sivesind_changing_2017}{Sivesind 2017}).

The question arises whether adults and children justify unequal access
to welfare services based on market justice principles. Influenced by
theories of policy feedback, which suggest that social welfare policies
can reinforce (positive feedback) or undermine (negative feedback)
previous policy trajectories
(\citeproc{ref-fernandez_positive_2013}{Fernandez and Jaime-Castillo
2013}; \citeproc{ref-pierson_increasing_2000}{Pierson 2000};
\citeproc{ref-weaver_paths_2010}{Weaver 2010}), citizens' beliefs about
market justice are likely also shaped by the institutional and social
contexts they encounter. Indeed, the justification of inequality in
access to essential services like education and health, based on one's
ability to pay, shows significant variation across countries, as
demonstrated by international surveys, although they do not usually
include children in their samples. For instance, the 2019 International
Social Survey Program (ISSP) provides insights into how adults perceive
inequality in accessing welfare services. Figure~\ref{fig-issp}
illustrates the variation in agreement levels regarding whether it is
just for individuals with higher incomes to purchase better education or
healthcare. Notably, in 18 of the 29 countries analyzed, there is a
greater justification for market inequality in healthcare access than in
education. Nevertheless, the general sentiment typically ranges from
`somewhat unjust' to 'neither just nor unjust, with mixed feelings.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-issp-1.pdf}

}

\caption{\label{fig-issp}Average of market justice preferences by
country}

\end{figure}%

While contemporary policy developments have increasingly embraced the
marketization of social welfare in areas such as education, pensions,
and health services, questions about public acceptance of these changes
persist. Lindh (\citeproc{ref-lindh_public_2015}{2015}) analysis of ISSP
2009 data from 17 OECD countries reveals a general lack of support for
market-based distribution of social services, suggesting widespread
disapproval of market stratification of essential services. This finding
is corroborated by Soler-Martínez, García-Sánchez, and Willis
(\citeproc{ref-soler-martinez_concerns_2023}{2023}) research from
Latinobarómetro 2020 across 18 Latin American countries, where concerns
about health and education access predominated over income inequality.
These results indicate that reforms toward welfare marketization are
typically driven by elite political decisions rather than grassroots
demand.

Despite high-income inequality and limited social mobility in Latin
America, there is a prevalent belief that individuals are solely
responsible for their economic outcomes, a view that varies across the
region (\citeproc{ref-bucca_merit_2016}{Bucca 2016};
\citeproc{ref-chong_mystery_2008}{Chong and Ñopo 2008};
\citeproc{ref-torche_intergenerational_2014}{Torche 2014};
\citeproc{ref-salgado_inequality_2023}{Salgado and Castillo 2023}). The
reliance on private welfare providers and widespread user fees
(\citeproc{ref-molyneux_neoliberal_2008}{Molyneux 2008}) adds complexity
to this context. Yet, research on children's justification of
market-based inequalities in accessing welfare services remains limited,
especially in Latin America, highlighting a significant gap in
understanding how younger generations view market-based access to
welfare and whether these views are associated with their meritocratic
beliefs.

@@

Empirical research on meritocracy has increased along with the
philosophical-normative discussion on meritocracy in recent years.
Particularly from a sociological perspective, meritocracy has been used
in research on social mobility to characterize societies with low
mobility that threaten the meritocratic ideal
(\citeproc{ref-goldthorpe_myth_2003}{Goldthorpe 2003}). More recently,
sociology and social psychology research has attended to the subjective
aspects vis-a-vis beliefs in meritocracy. The label of beliefs in this
realm covers a series of areas, such as attitudes, perceptions, and
preferences (Castillo et al), whereby most of the link this subjective
dimensions to individual socio-structural factors and context-level
determinants. For instance, some studies have analyzed how those with
greater privileges believe more in meritocracy (Reynolds \& Chan 2014),
how greater economic inequality increases meritocratic beliefs
(\citeproc{ref-mijs_paradox_2021}{Mijs 2021}), and how larger inequality
affects meritocratic beliefs
(\citeproc{ref-morris_representing_2022}{Morris et al. 2022}). Based on
these findings, a research agenda has been reinforced on the
legitimizing role of meritocracy, in line with previous studies using
the concept of a just world (Lerner, Dalbert) and the theory of system
justification (Jost \& Major).

How do meritocratic beliefs legitimize inequalities? Empirical studies
have used experiments and surveys to address this question. For
instance, the evidence suggests that just world beliefs correlate
negatively with support for redistributive compensation systems
(\citeproc{ref-frank_performance_2015}{Frank, Wertenbroch, and Maddux
2015}). Conversely, individuals tend to support redistribution when they
believe that the disadvantaged lack the opportunities to succeed
(\citeproc{ref-evans_strong_2018}{Evans and Kelley 2018}). Almås,
Cappelen, and Tungodden (\citeproc{ref-almas_cutthroat_2020}{2020})
found that in a relatively unequal society (the United States), the
highly educated accept inequality significantly more than the less
educated because they perceive inequality as justifiable owing to
differences in productivity (i.e., merit), whereas in a relatively equal
society (Norway), the less educated accept inequality more, but not
significantly more than the highly educated because meritocratic values
are less prevalent. Barr and Miller
(\citeproc{ref-barr_effect_2020}{2020}) also addressed this triple
interaction between the level of inequality in a society, the individual
level of education, and the perceived origin of the disparity (either by
luck or effort) to determine the extent to which inequality is accepted.
They found that the interaction's mechanism varies depending on the
compared societies. Finally, García-Sánchez et al.
(\citeproc{ref-garcia-sanchez_attitudes_2020}{2020}), using data for 41
countries from the International Social Survey Programme (ISSP), found
that the perceived size of the income gap correlated positively with
support for progressive taxation. Still, this association was weaker
among those who endorsed meritocratic and equal opportunity beliefs. In
the same line, experimental research by Durante and Putterman
(\citeproc{ref-durante_preferences_2009}{2009}) subjects support less
redistribution when the initial distribution is determined according to
task performance.

Research about meritocratic beliefs at school age is rather scarce,
leaving a wide research gap as schools are one of the primary
socialization institutions where achievement based on merit explains
success (\citeproc{ref-erivwo_meritocracy_2021}{Erivwo et al. 2021}).
Nevertheless, it is possible to find some initial works focused on the
area of distributive justice at school that are closely related to
meritocratic beliefs, as the ones by Resh and Sabbagh (quotes). Using
justice in grade obtention as a measure of distributive justice and
meritocracy (Sabbagh et al 2006), they find for instance that a larger
sense of distributive justice about grades is associated to higher
socio-economic status (\citeproc{ref-resh_sense_2010}{Resh 2010}), have
a positive effect on liberal democratic orientation and on trust in
people and in formal institutions (\citeproc{ref-resh_sense_2014}{Resh
and Sabbagh 2014}), and tend to refrain from violence and to engage to a
greater extent in extra-curricular school activity and community
volunteering (\citeproc{ref-resh_sense_2017}{Resh and Sabbagh 2017}).

@@

Research indicates various socialization practices at families and
schools during childhood and adolescence that impact dispositional and
behavioral tendencies concerning justifications of inequality in adult
life. The differences in economic understanding across age groups are
consistent with research on cognitive development
(\citeproc{ref-choudhury_social_2006}{Choudhury, Blakemore, and Charman
2006}). Adolescents with mature socio-cognitive abilities tend to
express stronger preferences for fairness than infants and children
(\citeproc{ref-wynn_not_2018}{Wynn et al. 2018}). As children grow
older, they become more likely to behave fairly, with their
early-emerging strict egalitarianism being replaced by an increasing
endorsement of fairness principles and engagement in collaborative
activities (\citeproc{ref-huppert_development_2019}{Huppert et al.
2019}; \citeproc{ref-mcauliffe_developmental_2017}{McAuliffe et al.
2017}). In these activities, their fairness views consider individual
contributions, merits, and circumstances
(\citeproc{ref-almas_fairness_2010}{Almås et al. 2010};
\citeproc{ref-huppert_development_2019}{Huppert et al. 2019};
\citeproc{ref-sigelman_development_1991}{Sigelman and Waitzman 1991}).
Engelmann and Tomasello (\citeproc{ref-engelmann_children_2019}{2019})
claim that children's sense of fairness emerges at three years old, and
we can observe it in collaborative activities, where they accept
inequality if the procedure gives everyone an equal chance. Therefore,
children at this age respond to unequal distributions based on
interpersonal concerns, as they already demand equal respect. In any
case, between 3 and 8 years of age, inequitable and anti-meritorious
allocations are evaluated more negatively, but equitable and meritorious
allocations are not evaluated more positively
(\citeproc{ref-elenbaas_unfairness_2019}{Elenbaas 2019}).

Some research shows that the social environment in which children
develop, such as family and school, is associated with their prosocial
behaviors by playing an essential role in the transmission of equity
norms (\citeproc{ref-schunk_fairness_2023}{Schunk and Zipperle 2023};
\citeproc{ref-kosse_prosociality_2020}{Kosse and Tincani 2020}). In
fact, schools contribute to institutionalizing and reproducing
inequality by promoting values, norms, practices, and languages familiar
to higher-class families because the dominant group's culture shapes
educational institutions
(\citeproc{ref-bourdieu_reproduction_1990}{Bourdieu and Passeron 1990}).
Middle- and upper-class students are better equipped to face academic
challenges and are more familiar with academic expectations
(\citeproc{ref-mikus_children_2020}{Mikus, Tieben, and Schober 2020}).
Such familiarity represents cultural capital in educational contexts
because higher-status students come to school ready to meet these
expectations and reap the benefits (\citeproc{ref-jack_no_2016}{Jack
2016}; \citeproc{ref-khan_privilege_2011}{Khan 2011}). Conversely,
lower-status children lacking cultural capital must catch up while
experiencing inequitable comparisons
(\citeproc{ref-goudeau_hidden_2017}{Goudeau and Croizet 2017}).
Additionally, academic achievement is treated as the outcome of
dispositional factors (e.g., pupils' efforts and talents or lack of
them) rather than the result of differential access to critical
resources. Due to the meritocratic frame schools encourage, both low-
and high-status individuals believe that students' success or failure is
not due to their family background but rather to differences in efforts
and talents (\citeproc{ref-darnon_where_2018}{Darnon et al. 2018}). In
this sense, we believe that the perception of meritocracy can influence
students' judgments about market justice preferences. Furthermore, we
believe there is a difference between students' perceptions of
meritocracy based on their own experience in school and what they
perceive in society at large. Consistently, our first two hypotheses
are:

@@

\begin{figure}

\centering{

\includegraphics{input/img/hypotheses.png}

}

\caption{\label{fig-hypotheses}Summary of hypotheses}

\end{figure}%

@@

Table~\ref{tbl-desc-dependientes} shows the items used, their response
categories, and their frequencies.
